---
title: "PredictionAssignment"
author: "Uwe Draeger"
date: "3/26/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)

```

## Executive Summary

Lorem ipsunt ...


## Data import

```{r cars}

```

Final goal is to predict the manner in which the exercise was done in the test data set. 




## Exploration

An analysis of the test data shows that some variables are missing completely. It may not be sensible to construct a prediction model that relies on data not present when finally testing the model. This reasoning provides a first selection criterion to reduce the variables in the training data set. 

Result is tibble predictors ... 


The training data contains information on six different users with at least 2,600 observations. 
The number of observations in each class of execution (classe) is large, with a minimum of almost 500 obervation per bucket.

This might allow to build a separate model for each user and average them to end up with the average 

```{r users}

table(pml_training$user_name)
table(pml_training$user_name, pml_training$classe)

```



## Cross validation

As there is a independent test data, there is no need to set aside observations for final testing. However, it seems prudent to create a validation set to enable some estimation of expected prediction error.

One might also consider 

## Model building

Type of final model: Classification - has the exercise been performed as specified or what type of mistake has been made.




```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
